Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-08 index.qmd
section:introduction,issues,structural; process:write

<!-- recording 1 -->
<!-- Part 1; 00:00:00 - 00:09:10 -->

<!-- recording 2 -->
<!-- Part 2; 00:00:00 - 00:17:30 -->

<!-- 1. What does CJ do? -->
In *comparative judgment* (CJ) studies, judges assess the presence of a trait or competence in different stimuli by performing pairwise comparisons [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a]. The comparison produces a dichotomous outcome, indicating which stimulus is perceived to possess a higher trait level. After multiple rounds of pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] to process the resulting outcomes and estimate scores that reflect the underlying trait of interest. This method has been successfully employed in assessing the quality of written texts, where quality describes the underlying trait of interest and the texts serve as the stimuli [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

<!-- Specifically, CJ has been used to evaluate the quality of written texts [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023] and to assess the quality of speech [@Boonen_et_al_2020]. Additionally, it has been applied to evaluate various competencies, including mathematical problem-solving skills [@Jones_2015], engineering design skills [@Bartholomew_et_al_2018], conceptual understanding in algebra [@Jones_et_al_2019], statistical and English knowdlege [@Marshall_et_al_2020], and STEM knowledge and skills [@Bartholomew_et_al_2020]. -->

<!-- 2. Where are we now with DCJ? -->
Numerous studies have documented the effectiveness of CJ in assessing various traits and competencies over the past decade. These studies have emphasized three aspects of the method's effectiveness: its reliability, validity, and practical applicability. Research on reliability indicates that CJ requires a relatively small number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to produce trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023]. Furthermore, evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021]. Additionally, research on validity suggests that scores generated by CJ can accurately represent the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2018; @Bouwer_et_al_2023]. Finally, research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Jones_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].

@@

<!-- 3. But what are the problems of DCJ as a method? -->
Nevertheless, despite the growing number of CJ studies, the literature's unsystematic and fragmented research approaches have left several critical issues unaddressed. This research primarily focuses on three: the apparent disconnect between CJ's measurement and structural model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in CJ's measurement model, and the unclear role of comparison algorithms on the method's reliability and validity. The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aims to address all three concerns simultaneously.

@@

<!-- 1. The BTL model is CJ's measurement model -->
In a typical CJ study, the BTL model serves as the measurement model for CJ [@Andrich_1978; @Bramley_2008]. A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010]. For example, when evaluating text quality, the BTL model uses the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of the texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

<!-- 3. But what is the problem with this? -->
However, the statistical literature advises caution when using estimated scores to conduct additional analyses or hypotheses tests. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty can introduce bias into the analysis and reduce the precision of hypothesis tests. Notably, the direction and magnitude of this bias are often unpredictable; results may be attenuated, exaggerated, or remain unaffected, depending on the uncertainty in the scores and the actual effects being tested [@Kline_et_al_2023, p. 25; @Hoyle_et_al_2023, p. 137]. Furthermore, reduced precision in hypothesis tests weakens their statistical power, ultimately increasing the likelihood of committing type-I or type-II errors [@McElreath_2020].

@@

<!-- 4. So, what is the solution?  -->
To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Kline_et_al_2023; @Hoyle_et_al_2023, p. 138] and Item Response Theory (IRT) [@deAyala_2009; @Fox_2010; @vanderLinden_et_al_2017] recommend conducting these analyses and tests within a structural model. A structural model specifies how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010]. This approach allows analyses that can account for both the scores and their uncertainties simultaneously, rather than treating them as separate elements. Therefore, an integrated approach that combines CJ’s measurement and structural models can offer significant advantages.

@@

@@

<!--   [Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015]]{style="color:red;"}. -->

@@

# Theory {#sec-theory}
<!-- recording 1 -->
<!-- Part 2; 00:14:30 - 00:21:00 -->

@@

@@

>>>>

<!-- recording: Sven 24.10.04; time: 00:00:00 - 00:09:10 -->
<!-- recording: Sven 24.10.04; time: 00:14:30 - 00:21:00 -->
<!-- recording: Sven and Tine 24.10.18; time: 00:04:05 - 00:08:30 -->
<!-- recording: Sven and Tine 24.10.25; time: 00:00:00 - 00:17:00 -->

<!-- 1. What does CJ do? -->
In *comparative judgment* (CJ) studies, judges assess a specific trait or competence across various stimuli by performing pairwise comparisons [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a]. Each comparison produces a dichotomous outcome, indicating which stimulus is perceived to manifest a higher trait level. For example, when assessing text quality, judges compare pairs of written texts (the stimuli) to determine the relative quality each text exhibit (the trait) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

<!-- and spoken language [@Boonen_et_al_2020]. Additionally, it has been applied to evaluate various competencies, including mathematical problem-solving skills [@Jones_2015], engineering design skills [@Bartholomew_et_al_2018], conceptual understanding in algebra [@Jones_et_al_2019], statistical and English knowledge [@Marshall_et_al_2020], and STEM knowledge and skills [@Bartholomew_et_al_2020]. -->

<!-- 2. Where are we now with DCJ? -->
Numerous studies have documented the effectiveness of CJ in assessing traits and competencies over the past decade. These studies have emphasized three aspects of the method's effectiveness: its reliability, validity, and practical applicability. Research on reliability indicates that CJ requires a relatively small number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to produce trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023]. Furthermore, evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021]. On the other hand, research on validity suggests that scores generated by CJ can accurately represent the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2018; @Bouwer_et_al_2023]. Finally, research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Jones_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].

@@

<!-- 3. But what are the problems of DCJ as a method? -->
Nevertheless, despite the growing number of CJ studies, unsystematic and fragmented research approaches in the literature have left several critical issues unaddressed. This research primarily focuses on three: the apparent disconnect between CJ's measurement and structural model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in CJ's measurement model, and the unclear role of comparison algorithms on the method's reliability and validity. The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aims to address all three concerns simultaneously.

@@

<!-- recording: Sven 24.10.04; time: 00:14:30 - 00:21:00 -->
<!-- recording: Sven and Tine 24.10.18; time: 00:08:30 - 00:14:55 -->
<!-- recording: Sven and Tine 24.10.25; time: 00:17:00 - 00:31:00 -->

<!-- 1. The BTL model is CJ's measurement model -->
In a typical CJ study, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] serves as the measurement model for CJ [@Andrich_1978; @Bramley_2008]. A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010]. For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of the texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

<!-- 3. But what is the problem with this? -->
However, the statistical literature advises caution when using estimated scores to conduct additional analyses or hypotheses tests. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty can introduce bias into the analysis and reduce the precision of hypothesis tests. Notably, the direction and magnitude of the bias are often unpredictable; results may be attenuated, exaggerated, or remain unaffected, depending on the amount of uncertainty present in the scores and the actual effects being tested [@Kline_et_al_2023, pp. 25; @Hoyle_et_al_2023, pp. 137]. Furthermore, reduced precision in hypothesis tests weakens their statistical power, ultimately increasing the likelihood of committing type-I or type-II errors [@McElreath_2020].

@@

<!-- 4. So, what is the solution?  -->
To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp. 138] and Item Response Theory (IRT) [@Fox_2010, chap. 6; @vanderLinden_et_al_2017_I, chap. 24] recommend conducting these analyses and tests within a structural model. A structural model specifies how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010]. This approach allows analyses that can account for both the scores and their uncertainties simultaneously, rather than treating them as separate elements. Therefore, an integrated approach that combines CJ’s measurement and structural models can offer significant advantages.

@@

<!-- recording: Sven and Tine 24.10.25; time: 00:31:00 - 00:52:20 -->

@@

<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015] -->

@@

# Theory {#sec-theory}

@@

<!-- general development -->
<!-- recording: Sven 24.10.04; time: 00:21:00 - 00:34:10 -->
<!-- recording: Sven 24.10.04; time: 00:41:00 - 00:46:00 -->
<!-- recording: Sven and Tine 24.10.18; time: 00:00:00 - 00:04:05 -->

<!-- comparison mechanism as missing mechanism -->
<!-- recording: Sven 24.10.04; time: 00:09:10 - 00:14:30 -->
<!-- recording: Sven 24.10.04; time: 00:34:10 - 00:41:00 -->

@@

<!-- recording: Sven 24.10.04; time: 00:46:00 - 01:01:00 -->

<!-- Using assumptions from IRT [@deAyala_2009, p. 20-21], we are assuming a multidimensional model, that violates the local independece assumption -->

<!-- [@deAyala_2009, p. 291-29] think about indeterminancies in MIRT. to solve metric indeterminancy you commonly fix a mu=0, and s=1, but for rotational indeterminancy, you need to define (maybe) orthogonality between units traits and judges bias -->
